---
title: "Intro_to_R"
author: "Kevin Healy"
date: "20/08/2023"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

## Intro to R intro

This is the first intro to R. The best way to learn R is to use it. The 
learning curve at the start is steep but once you get used to the basics of 
coding the rest is mainly googling.

## R is just a fancy calculator

You can add, multiply etc
```{r, calculator}
1+1 

#multiply
20*18

```

To print names use " "
```{r, text}
#print text
#It is important that you use " " for text

"Hello"
```

If you want R to remember something you need to save it as an object using 
the <- symbol. Do not use the equal symbol = for storing objects. Notice when 
you create an object it appears in the Global environment in the top right box.

```{r, objects}
#no. of oranges
oranges <- 10

#no. of apples
apples <- 5
```

If you want to print out the object simply type the name

```{r, oranges}
#if you want to print what is in the object
oranges
```


R, like any code, will read everything in the most literal way so any difference 
in spelling, upper and lower case or spaces between words will lead to an error.

Its hard to break R so feel free to make errors.

```{r, test, error = TRUE}

orange

Oranges

```


# Vectors, databases and other objects.

There are several classes of data, we will deal with Numeric, Character and 
Logic here. Numeric is numbers, Character is words and Logic is a special type
used to make arguments. We cal ask what classa data type is in R using class().

```{r, class type }

#numeric data
class(1)

#character data
class("hello")

#logical data
class(TRUE)

```

We can also store these data as vectors using c().

Lets say we say we observed the number of birds over 5 days, we could store the 
data like below.

```{r, vector}

no_birds <- c(2,5,10,24,12)
no_birds

```

Numbers are stored as "numeric" however, we can also store words as characters.
Lets say we recorded the weather as above.

```{r, vector cat}

weather <- c("rain","rain", "sun", "rain", "sun")
weather

```


Logical class data arises when we ask if something is true or false. For 
example, we can ask when is TRUE that an entrey in the vector weather is rain
using the below line.

```{r, vector logical}

weather == "rain"

```

Dataframes combine data into rows and column in the same way you would deal 
with them in excel

```{r, dataframe}

#data.frames
bird_watching <- data.frame(no_birds, weather)
bird_watching

```

We can query the rows and column of the dataframe using square brackets 
dataframe[row numbers, column numbers]

```{r, rows and cols}
#row 1 column 1 
bird_watching[1,1]

#row 3 column 2 
bird_watching[3,2]

#all of row 1
bird_watching[1,]

#all of column 1
bird_watching[,1]

```

We will cover this in more detail next week but we can also subset from 
data.frames using the name of the column.

```{r, names of cols}

#Can also use the name of the column
bird_watching[,"weather"]

```

We can also subset our data.frames using simple arguments. We can ask R which 
entries are equal to a given query. For example, for which rows was it raining.

```{r, raining}

#This asks which entries match "rain"
bird_watching[,"weather"] == "rain"

```

We can do these queries within the square brackets and R will return only 
the rows/column for which the query is TRUE


```{r, True raining}

#The data.frame will only return results that are TRUE
bird_watching[bird_watching[,"weather"] == "rain",]

```

We can also ask to return values when it is not TRUE that it is raining by 
using !=

```{r, False raining}

# !=  means not TRUE
bird_watching[,"weather"] != "rain"

#The dataframe will only return results that are TRUE
#but here we will get values for when it is TRUE that
#the value is not rain.
bird_watching[bird_watching[,"weather"] != "rain",]


```


There are many ways to subset data. However, while you might see it in some 
intro courses do not use the attach() function. This automatically creates a 
object for the each column. This can lead to overwriting previous objects.

```{r, subseting}

#There are many ways to subset data the same way

bird_watching[,"weather"]

bird_watching[,1]

bird_watching$weather

bird_watching[,]$weather

```


# Functions

One of the most powerful reasons to use R is the use of functions. A function
is some process where we do something to some set of data. For example, a 
function could be as simple as adding 1 to a list of numbers. However, we often 
want to do something more interesting. For example, we might want to find the 
mean of some set of numbers.


```{r, functions}

#create some list of numbers
input_data <- c(20,10,15)

mean(input_data)

```

If you want to know more about a function you can type ? before the function
and a help file will appear.

```{r, functions help}

?mean

```






